From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-08-06 WOLFHPC15/article/ms_msc.tex
wolfhpc fix all parts helene 06_08

\begin{figure}[t]
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++,numbers=left]
program ::= "mesh:" meshid 
            "mesh entities:" listmeshent
            "computation domains:" 
                       listcompdom
            "independent:"
                       listinde
            "data:" listdata
            "time:" iteration
            "computations:" listcomp
listmeshent ::= meshent listmeshent
             |  meshent
listcompdom ::= compdom listcompdom
             |  compdom
compdom ::= compdomid "in" listmeshent
listinde ::= inde listinde
          |  inde
inde ::= compdomid "and" compdomid
listdata ::= data listdata
          |  data
data ::= dataid "," meshent
iteration ::= num
listcomp ::= comp listcomp
          |  comp
comp ::= dataid "[" compdomid "]=" compid "(" 
            listdataread ")"
listdataread ::= dataread listdataread
              |  dataread
dataread ::= dataid "[" neighborid "]"
          |  dataid
\end{lstlisting}
\caption{Grammar of the MS language}
\label{fig:grammar}
\end{figure}

As mentioned in the previous section, the Multi Stencil Language (MSL) is an agnostic descriptive language for multi-stencil simulations. Six main sections are required in a MSL description that match the six elements of the formal definition; they describe: \textbf{1.}\,the mesh, \textbf{2.}\,the mesh entities, \textbf{3.}\,the computation domains with their inter-dependencies, \textbf{4.}\,the data, \textbf{5.}\,the time iterations, and finally \textbf{6.} the computations.
Figure~\ref{fig:grammar} shows the grammar of MSL. Lines 1 to 9 define what is a MSL program with the different parts mentioned above. The remaining of this section describes those different parts of a MSL program and presents an example.
We can note that, as MSL is a descriptive language, almost all terminals \texttt{meshid}, \texttt{meshdom}, \texttt{compdomid}, \texttt{dataid}, \texttt{compid} and \texttt{neighborid}, are string identifiers. Only \texttt{num} is an integer terminal.

\subsubsection*{Mesh and Mesh Entities}
As illustrated in the first line of the grammar in Figure~\ref{fig:grammar}, a mesh is simply defined by a string identifier.
Kind of mesh entities are described as a list of identifiers (lines 2, 10 and 11 in Figure~\ref{fig:grammar}).
Figure~\ref{fig:mslex} presents an example of a mesh named \texttt{cart} used to represent a Cartesian mesh, and three kind of mesh entities: \texttt{cell}, \texttt{edgex}, and \texttt{edgey}.

\subsubsection*{Computation Domains and Their Dependencies}
A computation domain is a subset of a given kind of entities from a mesh used during one or more computations.
For each computation domain, the kind of mesh entities on which it is mapped is specified (lines 3-4 and 12-14 on Figure~\ref{fig:grammar}).
By default two computation domains of the same entities are handled as intersecting except when an explicit independence relation is mentioned in the \emph{independent} section of the language (lines 5-6 and 15-17 in Figure~\ref{fig:grammar}).
On lines 4 to 6 of Figure~\ref{fig:mslex}, computation domains are defined for each kind of mesh entities: \texttt{allcell}, \texttt{alledgex} and \texttt{alledgey}; they are used to represent the whole domain.
On lines 7 and 8 two other domains of \texttt{edgex} are defined: \texttt{part1edgex} and \texttt{part2edgex}; they are used to represent the boundaries of the domain.
On line 10 the independence relation between \texttt{part1edgex} and \texttt{part2edgex} is specified.

@@

\subsubsection*{Computation Description}
The last part of the language contains the specifications of stencil kernels and local computations as defined in Section~\ref{sect:multistencil}.
This description does not contain a direct expression of the numerical expressions.
Instead, the term $exp$ of Equations~\ref{eq:st} and~\ref{eq:loc} simply corresponds to an identifier that references an implementation in an external language  (\cf Section~\ref{sect:component}).
The specification starts by the identifier of the computed data element ($w$ in the formal description) with the computation domain between brackets ($d$).
Then, after the equal sign, the kernel identifier ($exp$) is specified.
Finally, between the parenthesis is the list of identifiers of the data elements read by the computation with their stencil shape between brackets ($R$).
If the computation is local, no brackets appear ($R_l$). In the example of Figure~\ref{fig:mslex}, nine computations are described. On line 24, the data element \texttt{b} is computed on the domain \texttt{allcell} by the kernel \texttt{c\_0} which read data \texttt{a} without neighborhood shape.

@@

%----------------------------------------
\subsection{MSC Parallelization}
%----------------------------------------
The MSC parallelization is a subpart of the overall compiler MSCAC.
It makes use of the ordered list of computations $\Gamma$, which can directly be extracted from the parser, to build a parallel representation of the computations of the overall multi-stencil program.
This parallelization phase is divided in five different steps to transform $\Gamma$ to a series-parallel tree decomposition~\cite{Valdes:1979:RSP:800135.804393}.
Data parallelism is handled in the first two steps, and task parallelism in the following three steps.
This section gives an overview of these five steps , but they are more formally defined with their associated algorithms in a research report~\cite{????}.

@@

\subsubsection*{The Synchronized Ordered List of Computations: $\Gamma_{sync}$}
In a data parallelization, data is split among processors and the same program is applied on each sub-part of the data.
This does however require additional synchronizations between processors.
For example, as a stencil kernel accesses neighbor values, values computed by another processor are needed and must be synchronized. 
The required synchronizations can be automatically computed from the ordered list of computations $\Gamma$.
A synchronization is needed each time a data read by a stencil computation has been written by a previous computation. In such a case, a computation is added before the stencil computation. This \emph{synchronization computation} reads the data to synchronize, and write the same data. The computation domain of such a synchronization computation encompasses the whole mesh entities on which the data is declared. As a result $\Gamma$ is transformed to a synchronized ordered list of computations $\Gamma_{sync}$.
For example, in Figure~\ref{fig:mslex}, the stencil computation $c_1$ reads the data element $b$ which has been written by $c_0$. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed into $[c_0,sync_1,c_1]$ in $\Gamma_{sync}$. The new computation $sync_1$ reads and writes $b$ and is applied on the computation domain of $c_1$. As a result, a dependency is kept between $c_0$ and $sync_1$ and between $sync_1$, and $c_1$. The same transformation is performed for all stencils: $c_4$ and $c_8$. Thus, $\Gamma_{sync} = [c_0,sync_1,c_1,c_2,c_3,sync_4,c_4,c_5,c_6,c_7,sync_8,c_8]$.
\todo[inline]{JB: "is applied on the computation domain of $c_1$" vraiment, c'est pas ce que tu dis juste avant ...}

@@

\subsubsection*{The Minimal Series-Parallel Graph: $\Gamma_{msp}$}
Once a dependency graph is built, many solutions can be used to build a parallel application, as for example dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. In this work a static scheduling of the dependency graph is built. To do so the dependency graph is transformed to a minimal series-parallel graph~\cite{Valdes:1979:RSP:800135.804393}. As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a DAG is a minimal series-parallel graph if and only if a forbidden shape, called \emph{N-shape} illustrated in Figure~\ref{fig:n}, is not found in the DAG. Thus, to transform $\Gamma_{dep}$ to the minimal series-parallel graph $\Gamma_{msp}$, an algorithm~\cite{Mitchell:2004:CMV:1082101.1082117} is applied to remove all \emph{N-shapes} by an over-constraint, as illustrated in Figure~\ref{fig:over}.

\todo[inline]{JB: dans le DAG il y a une dependance de c2 a sync4, le graphe serie parallele les execute en parallele ...}

@@

  \path[->]
    (bc0) edge [dotted] node {} (c0)
    (bc2) edge [dotted] node {} (c2)
    (c0) edge node {} (c1)
          edge [dashed] node [swap] {} (c3)
    (c2) edge node {} (c1)
        edge node {} (c3);
  \end{tikzpicture}
}
  \caption{Forbidden $N$ subgraph shape for a DAG to be minimal series-parallel.}
  \label{fig:forbidden}
\end{center}
\end{figure}

@@

>>>>

\begin{figure}[t]
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++,numbers=left]
program ::= "mesh:" meshid 
            "mesh entities:" listmeshent
            "computation domains:" 
                       listcompdom
            "independent:"
                       listinde
            "data:" listdata
            "time:" iteration
            "computations:" listcomp
listmeshent ::= meshent listmeshent
             |  meshent
listcompdom ::= compdom listcompdom
             |  compdom
compdom ::= compdomid "in" listmeshent
listinde ::= inde listinde
          |  inde
inde ::= compdomid "and" compdomid
listdata ::= data listdata
          |  data
data ::= dataid "," meshent
iteration ::= num
listcomp ::= comp listcomp
          |  comp
comp ::= dataid "[" compdomid "]=" compid "(" 
            listdataread ")"
listdataread ::= dataread listdataread
              |  dataread
dataread ::= dataid "[" neighborid "]"
          |  dataid
\end{lstlisting}
\caption{Grammar of the Multi-Stencil Language}
\label{fig:grammar}
\end{figure}

As mentioned in the previous section, the Multi Stencil Language (MSL) is an agnostic descriptive language for multi-stencil simulations. Six main sections are required in a MSL description that match the six elements of the formal definition; they describe: \textbf{1.}\,the mesh, \textbf{2.}\,the mesh entities, \textbf{3.}\,the computation domains with their inter-dependencies, \textbf{4.}\,the data, \textbf{5.}\,the time iterations, and finally \textbf{6.} the computations.
Figure~\ref{fig:grammar} shows the grammar of MSL. Lines 1 to 9 define what is a MSL program with the different parts mentioned above. The remaining of this section describes those different parts of a MSL program and presents an example.
One can notice that almost all terminals \texttt{meshid}, \texttt{meshdom}, \texttt{compdomid}, \texttt{dataid}, \texttt{compid} and \texttt{neighborid}, are string identifiers. Only \texttt{num} is an integer terminal.

\subsubsection*{Mesh and Mesh Entities}
As illustrated in the first line of the grammar in Figure~\ref{fig:grammar}, a mesh is simply defined by a string identifier.
The different kind of mesh entities are described as a list of identifiers (lines 2, 10 and 11 in Figure~\ref{fig:grammar}).
Figure~\ref{fig:mslex} presents an example of a mesh named \texttt{cart} used to represent a Cartesian mesh, and three kind of mesh entities: \texttt{cell}, \texttt{edgex}, and \texttt{edgey}.

\subsubsection*{Computation Domains and Their Dependencies}
A computation domain is a subset of a given kind of mesh entities, used during one or more computations.
For each computation domain, the kind of mesh entities on which it is mapped is specified (lines 3-4 and 12-14 on Figure~\ref{fig:grammar}).
By default, two computation domains of the same entities are handled as intersecting except when an explicit independence relation is mentioned in the \emph{independent} section of the language (lines 5-6 and 15-17 in Figure~\ref{fig:grammar}).
On lines 4 to 6 of Figure~\ref{fig:mslex}, computation domains are defined for each kind of mesh entities: \texttt{allcell}, \texttt{alledgex} and \texttt{alledgey}; they are used to represent the whole domain.
On lines 7 and 8 two other domains of \texttt{edgex} are defined: \texttt{part1edgex} and \texttt{part2edgex}; they are used to represent the boundaries of the domain.
On line 10 the independence relation between \texttt{part1edgex} and \texttt{part2edgex} is specified.

@@

\subsubsection*{Computation Description}
The last part of the language contains the specifications of stencil and local kernels as defined in Section~\ref{sect:multistencil}.
This description does not contain a direct expression of the numerical expressions $exp$.
Instead, the term $exp$ of Equations~\ref{eq:st} and~\ref{eq:loc} simply corresponds to an identifier that references an implementation in an external language  (\cf Section~\ref{sect:component}).
The specification starts by the identifier of the computed data element ($w$ in the formal description) with the computation domain between brackets ($d$).
Then, after the equal sign, the kernel identifier ($exp$) is specified.
Finally, between the parenthesis is the list of identifiers of the data elements read by the computation with their stencil shape between brackets ($R$).
If the computation is local, no brackets appear ($R_l$). In the example of Figure~\ref{fig:mslex}, nine computations are described. On line 24, the data element \texttt{b} is computed on the domain \texttt{allcell} by the kernel expression \texttt{c\_0} which read data \texttt{a} without neighborhood shape.

@@

%----------------------------------------
\subsection{MSC Parallelization}
%----------------------------------------
The MSC parallelization is a subpart of the overall compiler MSCAC.
It makes use of the ordered list of computations $\Gamma$, which can directly be extracted from the parser, to build a parallel representation of the computations of the overall multi-stencil program.
This parallelization phase is divided in five different steps to transform $\Gamma$ to a series-parallel tree decomposition~\cite{Valdes:1979:RSP:800135.804393}.
%Data parallelism is handled in the first two steps, and task parallelism in the following three steps.
This section gives an overview of these five steps , but they are more formally defined with their associated algorithms in a research report~\cite{????}.

@@

\subsubsection*{The Synchronized Ordered List of Computations: $\Gamma_{sync}$}
In a data parallelization, data is split among processors and the same program is applied on each sub-part of the data.
However, this parallelization technique requires additional synchronizations between processors.
For example, as a stencil kernel accesses neighbor values, values computed by another processor are needed and must be synchronized. 
The required synchronizations can be automatically computed from the ordered list of computations $\Gamma$.
A synchronization is needed each time a data read by a stencil computation has been written by a previous computation. In such a case, a computation is added before the stencil computation. This \emph{synchronization computation} reads the data to synchronize, and write the same data. The computation domain of such a synchronization computation encompasses the whole mesh entities on which the data is declared. As a result, $\Gamma$ is transformed to a synchronized ordered list of computations $\Gamma_{sync}$.
For example, in Figure~\ref{fig:mslex}, the stencil computation $c_1$ reads the data element $b$ which has been written by $c_0$. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed into $[c_0,sync_1,c_1]$ in $\Gamma_{sync}$. The new computation $sync_1$ reads and writes $b$ and is applied on \texttt{edgex}. As a result, a dependency is kept between $c_0$ and $sync_1$ and between $sync_1$, and $c_1$. The same transformation is performed for all stencils: $c_4$ and $c_8$. Thus, $\Gamma_{sync} = [c_0,sync_1,c_1,c_2,c_3,sync_4,c_4,c_5,c_6,c_7,sync_8,c_8]$.

@@

\subsubsection*{The Minimal Series-Parallel Graph: $\Gamma_{msp}$}
Once a dependency graph is built, many solutions can be used to build a parallel application, as for example dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. In this work a static scheduling of the dependency graph is built. To do so the dependency graph is transformed to a minimal series-parallel graph~\cite{Valdes:1979:RSP:800135.804393}. As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a DAG is a minimal series-parallel graph if and only if a forbidden shape, called \emph{N-shape} and illustrated in Figure~\ref{fig:n}, is not found in the DAG. Thus, to transform $\Gamma_{dep}$ to the minimal series-parallel graph $\Gamma_{msp}$, an algorithm~\cite{Mitchell:2004:CMV:1082101.1082117} is applied to remove all \emph{N-shapes} by an over-constraint, as illustrated in Figure~\ref{fig:over}.

@@

  \path[->]
    %(bc0) edge [dotted] node {} (c0)
    %(bc2) edge [dotted] node {} (c2)
    (c0) edge node {} (c1)
          edge [dashed] node [swap] {} (c3)
    (c2) edge node {} (c1)
        edge node {} (c3);
  \end{tikzpicture}
}
  \caption{Forbidden $N$ subgraph shape for a DAG to be minimal series-parallel.}
  \label{fig:forbidden}
\end{center}
\end{figure}

@@

\subsubsection*{Data parallelism optimizations: $\Gamma_{data}$}
Building the series-parallel tree decomposition $\Gamma_{tsp}$ is not only limited to the creation of a static scheduling of tasks. Actually, the tree can also be used to optimize the automatic generation of the data parallel structure of the simulation. In fact, creating a data parallelization of the simulation, it first seems natural to create the parallel structure directly from $\Gamma_{sync}$, as the program is parallelized from data decomposition and applying the same sequential algorithm on each processor. However, as a computation kernel in MSL writes a single data $w$ at a time (in order to entirely manage and order the dependencies between tasks), the same computation domain might be iterated more than once to compute different numerical expressions, while those expressions could be computed inside a single space loop. For this reason, an optimized and \emph{fusionned} ordered list of computation $\Gamma_{data}$ is built from $\Gamma_{tsp}$. Actually, in $\Gamma_{tsp}$, computations of a parallel subtree with a same computation domain can be computed in the same space loop, in any order. Moreover, in $\Gamma_{tsp}$, two consecutive computations of a sequence subtree with a same computation domain can be computed in the same space loop, if the order of computation of the sequence is kept.

### 2015-08-06 WOLFHPC15/article/mscac.tex
wolfhpc fix all parts helene 06_08

The proposed DSL, named MSL (\emph{Multi-Stencil Language}), is an agnostic descriptive language for multi-stencil simulations. Agnostic means that the description of a numerical simulation does not depend on implementation choices as, for example, the type of mesh (structured, unstructured), or its associated interfaces (how to define a stencil shape, etc.). Descriptive, on the other hand, means that the terminal states of the language grammar are (string) identifiers. As a result, MSL is not made to handle the expression of numerical computations, but only the description of a multi-stencil simulation, in a coarse-grain fashion. %Numerical computations are encapsulated in specific components that users have to develop. 
A MSL description is made of six main sections that match the six tuples elements of the definition of a multi-stencil program of Equation~\ref{eq:msp}.
\todo[inline]{JB: dans une grammaire, les identifiants sont toujours des chaines de caracteres, je ne comprends pas}

@@

The final phase of compilation, the dump, uses the static scheduling of the tasks, created from $\Gamma$, and additional informations on $T$, $\mathcal{M}$, $\Delta$, $\mathcal{E}$ and $\mathcal{D}$.
It generates a program in the back-end language which embeds the parallel pattern or skeleton of the overall simulation.
Various models could be used as back-end, such as OpenMP~\cite{660313} for example, however this paper adopts a novel approach where the dump targets a component-based runtime.
This approach makes it possible to leverage good properties of software component models such as increased code re-use, eased separation of concerns, and increased maintainability and productivity.
% JB: la phrase en dessous est pas tres heureuse ...
Those properties counter some known limitations of DSLs. % JB: was "parallel applications".
%where a component represents a clear and independent functionality of the application, have first been proposed in software engineering and have proved many times good properties for

Thanks to the software components based approach, the generated code is made of five independent parts that loosely match the elements of the multi-stencil program definition.
These are the components implementing: \textbf{a)} the scheduling ($\Gamma$), \textbf{b)} the data structures ($\mathcal{M}$ and $\mathcal{E}$), \textbf{c)} their parallelization ($\Delta$ and $\mathcal{D}$), \textbf{d)} the time loop ($T$), and \textbf{e)} the kernels directly provided by the user.
This enables to change the technology used for each part while limiting the impact on the other parts as further discussed in Section~\ref{sect:component}.

>>>>

The proposed DSL, named MSL (\emph{Multi-Stencil Language}), is an agnostic descriptive language for multi-stencil simulations. Agnostic means that the description of a numerical simulation does not depend on implementation choices as, for example, the type of mesh (structured, unstructured), or its associated interfaces (how to define a stencil shape, etc.). Descriptive, on the other hand, means that MSL is not made to handle the expression of numerical computations, but only the description of a multi-stencil simulation, in a coarse-grain fashion. %Numerical computations are encapsulated in specific components that users have to develop. 
A MSL description is made of six main sections that match the six tuples elements of the definition of a multi-stencil program of Equation~\ref{eq:msp}.

@@

The final phase of compilation, the dump, uses the static scheduling of the tasks, created from $\Gamma$, and additional information on $T$, $\mathcal{M}$, $\Delta$, $\mathcal{E}$ and $\mathcal{D}$.
It generates a program in the back-end language which embeds the parallel pattern or skeleton of the overall simulation.
Various models could be used as back-end, such as OpenMP~\cite{660313} for example, however this paper adopts a novel approach where the dump targets a component-based runtime.
This approach makes it possible to leverage good properties of software component models such as increased code re-use, eased separation of concerns, and increased maintainability and productivity.
% JB: la phrase en dessous est pas tres heureuse ...
Those properties counter some known limitations of DSLs. % JB: was "parallel applications".
%where a component represents a clear and independent functionality of the application, have first been proposed in software engineering and have proved many times good properties for

Thanks to the software components based approach, the generated code is made of four independent parts that loosely match the elements of the multi-stencil program definition.
These are the components implementing: \textbf{a)} the computations scheduling ($\Gamma$), \textbf{b)} the data structures ($\mathcal{M}$ and $\mathcal{E}$), \textbf{c)} data of the simulation ($\Delta$ and $\mathcal{D}$), and \textbf{d)} the time loop ($T$). The actual implementation of the numerical expressions are provided by the user and used in the computations scheduling.
This component-based parallel structure enables to change the technology used for each part while limiting the impact on the other parts as further discussed in Section~\ref{sect:component}.

### 2015-08-06 WOLFHPC15/article/related.tex
wolfhpc fix all parts helene 06_08

Many specific solutions exist to ease parallel programming of numerical simulations. Solutions such as PETSc~\cite{petsc-efficient} and Trilinos~\cite{Trilinos-Overview} are specific libraries. They are made to solve many kinds of PDEs using well-known methods and solvers which are efficiently parallelized inside the library. Those libraries also include solvers for implicit numerical schemes, not addressed in this work. However, using those solutions the user has to use and follow existing tools. In other words, it is not possible for a mathematician or a physician to write his own numerical method if it differs from existing ones. Moreover as PETSc and Trilinos are libraries they do not propose a domain specific language to define computations. The purpose of those solutions are much more specific than our solution.
\todo[inline]{JB: j'utiliserais plutot le terme "framework" que "library"}

@@

Most stencil compilers handle the parallelization or the optimization of a single stencil kernel, considering that it represent the main computation time of numerical simulations. However, many real case numerical simulation are not composed of a single stencil kernel, as explained in Section~\ref{sect:multistencil}, but of more than one stencil computation, involving one or more stencil shapes, and additional local computations. Solutions like Pochoir or PATUS handle the parallelization or the optimization of a single stencil code, considering that stencil codes represent the main computation time of numerical simulations. Liszt~\cite{DeVito2011LDS} is the closest solution to our work as it seems to take into account the entire program and not only stencil kernels.

>>>>

Many specific solutions exist to ease parallel programming of numerical simulations. Solutions such as PETSc~\cite{petsc-efficient} and Trilinos~\cite{Trilinos-Overview} are specific frameworks. They are made to solve many kinds of PDEs using well-known methods and solvers which are efficiently parallelized inside the library. Those frameworks also include solvers for implicit numerical schemes, not addressed in this work. However, using those solutions the user has to use and follow existing tools. In other words, it is not possible for a mathematician or a physician to write his own numerical method if it differs from existing ones. Moreover as PETSc and Trilinos are frameworks they do not propose a domain specific language to define computations. The purpose of those solutions are much more specific than our solution.

@@

Most stencil compilers handle the parallelization or the optimization of a single stencil kernel, considering that it represent the main computation time of numerical simulations. However, many real case numerical simulation are not composed of a single stencil kernel, as explained in Section~\ref{sect:multistencil}, but of more than one stencil computation, involving one or more stencil shapes, and additional local computations. %Solutions like Pochoir or PATUS handle the parallelization or the optimization of a single stencil code, considering that stencil codes represent the main computation time of numerical simulations. 
Liszt~\cite{DeVito2011LDS} is the closest solution to our work as it seems to take into account the entire program and not only stencil kernels.
